Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-11 Manuscript/manuscript.Rmd
methods edits
intro edits

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome <!--may affect genomic regulation--> [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking; @dixon2012topological]. Likewise, evidence suggests that the boundaries that delineate TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes [@symmons2014functional]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d; @symmons2014functional]. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo]. Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the CCCTC-binding factor  (CTCF), has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]--> [??? What's the summary/conclusion/problem?]

Current methods used to predict [??? you never introduced prediction problem] TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These TAD boundary regions make up the minority class, while the non-TAD boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. A dataset is considered imbalanced if PI<0.5. [??? Not really. Poor metric choice.] <!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

<!--
[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).
-->

@@

## CLASSIFICATION

@@

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking]. Studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].  However, the growing number of publically available annotated genomic elements and there collinearity pose a significant problem for determining the most important epigenomic factors that contribute to TAD boudandary formation.

## PREVIOUS METHODS

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo]. Currently, classification models have been used to draw associations between these functional annotations and genomic regions that include TAD boundaries.

## PROBLEMS AND LIMITATIONS

## OUR APPROACH

##################
Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the CCCTC-binding factor  (CTCF), has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]--> [??? What's the summary/conclusion/problem?]

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These TAD boundary regions make up the minority class, while the non-TAD boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. A dataset is considered imbalanced if PI<0.5. [??? Not really. Poor metric choice.] <!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models. 
##################

@@

<!--[Paragraph about disconnect between Hi-C and epigenomic data]--> 
Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).

@@

## DEVELOPING FRAMEWORK TO DETERMINE OPTIMAL DATA LEVEL CHARACTERISTICS FOR TAD BOUNDARY PREDICTION

### CLASSIFICATION

@@

## USING OPTIMALLY BUILD CLASSIFIER TO PREDICT BOUNDARIES AT FINER RESOLUTION
